### 2024-04-20 MSRI-16-1989-79.Rmd
another big tree

::: {.rmenv #chunk-1.13 title="1.13 Remark" latex="{1.13 Remark}"}
\oldpage{98}
The treatment of crystalline structures is not satisfying.
In the definition of realisation systems, I have not included the data  given in [(1.7)](#chunk-1.7), for the want of verifying their existence in the case of Lie algebras of $\pi_1$ that interest us.
I have nevertheless included the data of $\phi_p$, despite its appearance as a bizarre addition, because the calculations in [§19](#section-19) give an interesting result.
:::

:::: {.rmenv #chunk-1.14 title="1.14" latex="{1.14}"}
Here is a variant of the statement of [(1.4)](#chunk-1.4).
This formulation, which is less elementary, highlights the role of $F_\infty$.

::: {.trmenv #Mp1 title="(M'1)" latex="{(M'1)}"}
To every algebraic closure $C$ of $\RR$ is attached, in a functorial way, $M_\BB(C)$.
:::

From [(M'1)](#Mp1) we deduce the data of [(M1)](#M1) and [(M7)](#M7) by setting $M_\BB\coloneqq M_\BB(\CC)$, and the taking $F_\infty$ induced by $z\mapsto\bar{z}\colon\CC\to\CC$.
For $M=\HH_\mot^i(X)$, we will have $M_\BB(C)=\HH^i(X(C),\QQ)$.

::: {.trmenv #Mp2 title="(M'2)" latex="{(M'2)}"}
The same as [(M2)](#M2).
:::

::: {.trmenv #Mp8 title="(M'8)" latex="{(M'8)}"}
The same as [(M8)](#M8).
:::

::: {.trmenv #Mp3 title="(M'3)" latex="{(M'3)}"}
An $\AAf$-sheaf $M_\AAf$ on $\Spec(\QQ)$.
:::

By "$\AAf$-sheaf" we mean the data, for all $\ell$, of a $\QQ_\ell$-sheaf $\mathcal{F}_\ell$, and, for almost all $\ell$, of a $\ZZ_\ell$-sheaf $\mathcal{F}_{\ZZ_\ell}\subset\mathcal{F}_\ell$ which generates $\mathcal{F}_\ell$: the germ of the system of the $\mathcal{F}_{\ZZ_\ell}$ is given.
On the spectrum of a field $k$, the data of $\mathcal{F}_\ell$ (resp. $\mathcal{F}_{\ZZ_\ell}$) is equivalent to that of, for every algebraic closure $\bar{k}$ of $k$, a $\QQ_\ell$-vector space $\mathcal{F}_\ell(\bar{k})$ (resp. a $\ZZ_\ell$-module $\mathcal{F}_{\ZZ_\ell}(\bar{k})$ of finite type), functorially in $\bar{k}$, and such that the action of $\Gal(\bar{k}/k)$ is continuous.
Note that $\mathcal{F}_\AAf(\bar{k})$ is the restricted product of the $\mathcal{F}_\ell(\bar{k})$ with respect to the $\mathcal{F}_{\ZZ_\ell}(\bar{k})$.

From [(M'3)](#Mp3), we deduce the data of [(M3)](#M3) and [(M9)](#M9) by setting $M_\AAf\coloneqq(M_\AAf)_\AAf(\bar{\QQ})$.
_Notation:_ we write $M_\ell$ (resp. $M_{\ZZ_\ell}$, $M_\AAf$) for $(M_\AAf)_\ell$ (resp. $(M_\AAf)_{\ZZ_\ell}$, $(M_\AAf)_\AAf$).

If $M=\HH_\mot^i(X)$, and $a$ is the morphism $X\to\Spec(\QQ)$, then $M_\ell=\RR^ia_*\QQ_\ell$, and $M_{\ZZ_\ell}$ is equal to the image of $\RR^ia_*\ZZ_\ell$ in $M_\ell$.
We have that $\RR^ia_*\ZZ_\ell(\bar{k})=\HH^i(X\otimes_\QQ\bar{k},\ZZ_\ell)$.

::: {.trmenv #Mp4 title="(M'4)" latex="{(M'4)}"}
For almost all primes $p$, an $F$-isocrystal $M_\crisp$ on $\FF_p$, i.e. a vector space $M_\crisp$ over $\QQ_p$ endowed with an automorphism $\phi_p$.
:::

::: {.trmenv #Mp5 title="(M'5)" latex="{(M'5)}"}
\oldpage{99}Comparison isomorphisms
$$
  \begin{aligned}
    \comp_{\DR,\BB}&\colon M_\BB(C)\otimes C \simto M_\DR\otimes C
  \\\comp_{\AAf,\BB}&\colon M_\BB(C)\otimes\AAf \simto M_\AAf(\bar{\QQ}_C)
  \\\comp_{\crisp,\DR}&\colon M_\DR\otimes\QQ_p \simto M_\crisp
  \end{aligned}
$$
that are functorial in $C$, where $\bar{\QQ}_C$ is the algebraic closure of $\QQ$ in $C$.
The first is equivalent to the data of $\comp_{\DR,\BB}$ as in [(M5)](#M5) satisfying [(AM2)](#AM2);
the second is equivalent to the data of $\comp_{\AAf,\BB}$ as in [(M5)](#M5) satisfying [(AM5)](#AM5).
:::

In [(M'3)](#Mp3), instead of giving the $M_\AAf$, we could have given only the $\QQ_\ell$-sheaves $M_\ell$, replacing $\comp_{\AAf,\BB}$ by the $\comp_{\ell,\BB}\colon M_B(C)\otimes\QQ_\ell\simto M_\ell(\bar{\QQ}_C)$ and requiring the existence of an integer lattice $L\subset M_B(\CC)$ such that the $\comp_{\ell,\BB}(L\otimes\ZZ_\ell)$ be stable under $\Gal(\bar{\QQ}/\QQ)$.
They define the $M_{\ZZ_\ell}$ of [(M'3)](#Mp3).

::: {.trmenv #Mp6 title="(M'6)" latex="{(M'6)}"}
$W$ is a filtration of the objects of [(M'1)](#Mp1) to [(M'4)](#Mp4), respected by the comparison isomorphisms.
:::

A definition essentially equivalent to \ref{minus-three-points-1.9} is then the following: a _realisation system_ is a system [(M'1)](#Mp1) to [(M'6)](#Mp6), and [(M'8)](#Mp8), satisfying axioms [(AM1)](#AM1), [(AM2)](#AM2), and [(AM4)](#AM4) (suitably modified in the evident way).

The data of [(M'1)](#Mp1) is equivalent to a sheaf of $\QQ$-vector spaces on the étale site of $\Spec\RR$.
From this point of view, $\comp_{\ell,\BB}$ is an isomorphism to $\Spec(\RR)_\et$ between the inverse image of $M_\ell$ and the $\QQ_\ell$-sheaf induced by $M_\BB$.

In the language of sheaves, [(AM3)](#AM3) implies that $M_\ell$ comes from a smooth $\QQ_\ell$-sheaf on $\Spec(\ZZ)\setminus S\setminus\{\ell\}$.
The language of sheaves makes it clear that, for all $p$, $M_\ell$ defines a $\QQ_\ell$-sheaf on $\Spec(\QQ_p)$ (cf. the analogous case of $\RR$ below).
The choice of a decomposition group is not required.
::::

::: {.rmenv #chunk-1.15 title="1.15 Variant" latex="{1.15 Variant}"}
Let $P$ be a finite set of prime numbers.
The category of *smooth realisation systems on $\Spec(\ZZ)\setminus P$* is defined as in \ref{minus-three-points-1.9}, taking $P$ to be the exceptional set $S$ in [(AM3)](#AM3), and replacing "almost all $p$" in [(M4)](#M4) and [(AM4)](#AM4) by "all $p\not\in P$".
This treatment of crystalline structures is not satisfying, cf. [(1.7)](#chunk-1.7).
The category \ref{minus-three-points-1.9} of realisation systems on $\Spec(\QQ)$ is the inductive limit of these categories for $P$ growing larger and larger.

@@

\oldpage{100}In the language of [(1.14)](#chunk-1.14), in [(M'3)](#Mp3) we need to give $M_\ell$ as a smooth $\QQ_\ell$-sheaf on $\Spec(\ZZ[1/\ell])\setminus P$ instead of on $\Spec(\QQ)$, and we need to modify [(M'4)](#Mp4) like [(M4)](#M4).

@@

We would like to have a notion of smooth motive on $S$ for more general base spaces than $\Spec(\ZZ)\setminus P$.
Our methods, where the Betti realisation plays a central role, require that $S_\QQ$ be dense in $S$.
We will outline a provisional definition of smooth realisation systems on $S$, for $S$ smooth over $\Spec(\ZZ)$.
The case where $S$ is open in the spectrum of the ring of integers of a number field can be dealt with using natural modifications of [(1.4)](#chunk-1.4) and [(1.15)](#chunk-1.15).
For $S$ finite and étale over an open of $\Spec(\ZZ)$, we can also reduce to [(1.15)](#chunk-1.15): see [(1.17)](#chunk-1.17).
:::

@@

::: {.trmenv #Mp3E title="(M'3)E" latex="{(M'3)\textsubscript{E}}"}
(cf. [(M3)](#M3), [(M9)](#M9)) An $\AAf$-sheaf $M_\AAf$ on $\Spec(E)$ (cf. [(1.14)](#chunk-1.14)).
:::

@@

By summing over $\sigma$ (resp. $v$), these give [(M'5)](#Mp5).

@@

We leave to the reader the task of translating axioms [(AM1)](#AM1), [(AM3)](#AM3), and [(AM4)](#AM4), which remain to be imposed, into this language (cf. [(1.14)](#chunk-1.14)).
::::

@@

- In [(AM3)](#AM3), we ask for each $\QQ_\ell$-sheaf $M_\ell$ induced by $M_\AAf$ to come from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$.
- In [(M'4)E](#Mp4E), for each residue field $k(v)$ of $S$, we ask for $M_\crisv$ over the field of fractions $K_v$ of Witt vectors over $k(v)$, endowed with a semi-linear $\phi_v$.
  The crystalline comparison isomorphism of [(M'5)E](#Mp5E) then becomes
  $$
    M_\DR\otimes E_v \simto M_\crisv\otimes_{K_v}E_v.
  $$
:::

@@

The axioms are modified as follows.
In [(AM1)](#AM1), we want a variation of mixed Hodge structures;
[(AM3)](#AM3) becomes: $M_\ell$ comes from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$;
for [(AM4)](#AM4), a condition is imposed for every closed point of $S$.

@@

- [(M1)](#M1) and [(M7)](#M7): $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $F_\infty=-1$.
- [(M'1)](#Mp1): $\ZZ(1)_\BB(C)=2\pi i\ZZ\subset C$.
- [(M2)](#M2) and [(M'2)](#Mp2): $\ZZ(1)_\DR=\QQ(1)_\DR=\QQ$.
- [(M3)](#M3) and [(M9)](#M9): $\ZZ(1)_{\ZZ/n}$ is the group $\mu_n\subset\CC^\times$ of $n$-th roots of unity.
  The transition morphisms are the $\mu_n\to\mu_m\colon x\mapsto x^{n/m}$ for $m\mid n$.
  The action of $\Gal(\bar{\QQ}/\QQ)$ is its action on the $\mu_n$.
  In the variant [(M'3)](#Mp3), we have $\ZZ(1)_{\ZZ/n}(\bar{k})=\mu_n(\bar{k})$.

  We thus have that $\ZZ(1)_\ell=\limproj\mu_{\ell^n}(\CC)$.
- [(M4)](#M4), [(M10)](#M10), and [(M'4)](#Mp4): $\ZZ(1)_\crisp=\QQ_p$ and $\phi_p=1/p$.
- [(M5)](#M5): The canonical comparison isomorphism $\comp_{\DR,\BB}$ is induced by the inclusions $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $\ZZ(1)_\DR=\QQ\subset\CC$, i.e.
  $$
    \ZZ(1)_\BB\otimes\CC \simto \CC \simfrom \ZZ(1)_\DR\otimes\CC.
  $$
  The isomorphism $\comp_{\ZZ/n,\BB}$ [(1.25)](#chunk-1.25) is induced by
  $$
    \exp(x/n)\colon \ZZ(1)_\BB \to \ZZ/n(1)\subset\CC^\times.
  $$
  The isomorphism $\comp_{\crisp,\DR}$ is induced by the inclusion $\QQ\subset\QQ_p$.
- [(M'5)](#Mp5): Replace $\CC$ by $C$ in [(M5)](#M5).
- [(M6)](#M6) and [(M8)](#M8): $\ZZ(1)$ is of pure weight $-2$, and $\ZZ(1)_\DR$ is of pure Hodge filtration $-1$: the Hodge type is $(-1,-1)$.
:::

@@

::: {.trmenv #M7P title="(M7)P" latex="{(M7)P}"}
\oldpage{108}The data of $F_\infty$ for $E$ is equivalent to that of an involution of $P_\BB$ that is compatible with that of $M_\BB$.
Axiom [(AM2)](#AM2) for $E$ is that this involution is induced by the $\CC$-antilinear involution of $P_{\BB,\CC}=\PP_{\DR,\CC}$ with fixed points $P_{\DR,\RR}$.
:::

::: {.trmenv #M8P title="(M8)P" latex="{(M8)P}"}
The data of the Hodge filtration of $E_\DR$ is equivalent to that of the intersection $F^0P_\DR$ of $F^0E_\DR$ over $P_\DR$, which is a torsor for $F^0M_\DR$: for $p>0$, $F^pE_\DR=F^pM_\DR$; for $p=0$, $F^0E_\DR$ is generated by $F^0P_\DR$; and for $p<0$, $F^pE_\DR$ is the sum of $F^0E_\DR$ and $F^pM_\DR$.
In other words, the data [(M8)](#M8) for $E_\DR$ is that of a reduction of the structure group of $P_\DR$, from $M_\DR$ to $F^0M_\DR$.

The data of [(M1)](#M1), [(M2)](#M2), $\comp_{\DR/\BB}$, and [(M8)](#M8) are equivalent to the data of a torsor $F^0P_\DR$ for $F^0M_\DR$ and, in the $M_\DR\otimes\CC=M_\BB\otimes\CC$ torsor $P_{\DR,\CC}$ which is induced by $F^0M_\DR\to M_\DR\otimes\CC$, a $\QQ$-structure $P_\BB\subset P_{\DR,\CC}$ that is an $M_\BB$ torsor.
:::

@@

We leave to the reader the task of translating axioms [(AM1)](#AM1) to [(AM5)](#AM5) for $P$.
We note only that [(AM1)](#AM1) and [(AM4)](#AM4) follow from the same axioms for $M$, and that [(AM2)](#AM2) and [(AM5)](#AM5) can be expressed at best as functoriality in $\CC$, cf. [(1.14)](#chunk-1.14).
::::

>>>>

::: {.rmenv #chunk-1.15 title="1.15 Variant" latex="{1.15 Variant}"}
Let $P$ be a finite set of prime numbers.
The category of *smooth realisation systems on $\Spec(\ZZ)\setminus P$* is defined as in \ref{minus-three-points-1.9}, taking $P$ to be the exceptional set $S$ in [(AM3)](am3), and replacing "almost all $p$" in [(M4)](m4) and [(AM4)](am4) by "all $p\not\in P$".
This treatment of crystalline structures is not satisfying, cf. [(1.7)](#chunk-1.7).
The category \ref{minus-three-points-1.9} of realisation systems on $\Spec(\QQ)$ is the inductive limit of these categories for $P$ growing larger and larger.

@@

\oldpage{100}
In the language of [(1.14)](#chunk-1.14), in [(M'3)](mp3) we need to give $M_\ell$ as a smooth $\QQ_\ell$-sheaf on $\Spec(\ZZ[1/\ell])\setminus P$ instead of on $\Spec(\QQ)$, and we need to modify [(M'4)](mp4) like [(M4)](m4).

@@

We would like to have a notion of smooth motive on $S$ for more general base spaces than $\Spec(\ZZ)\setminus P$.
Our methods, where the Betti realisation plays a central role, require that $S_\QQ$ be dense in $S$.
We will outline a provisional definition of smooth realisation systems on $S$, for $S$ smooth over $\Spec(\ZZ)$.
The case where $S$ is open in the spectrum of the ring of integers of a number field can be dealt with using natural modifications of \ref{minus-three-points-1.4} and [(1.15)](#chunk-1.15).
For $S$ finite and étale over an open of $\Spec(\ZZ)$, we can also reduce to [(1.15)](#chunk-1.15): see [(1.17)](#chunk-1.17).
:::

@@

::: {.trmenv #Mp3E title="(M'3)E" latex="{(M'3)\textsubscript{E}}"}
(cf. [(M3)](m3), [(M9)](m9)) An $\AAf$-sheaf $M_\AAf$ on $\Spec(E)$ (cf. [(1.14)](#chunk-1.14)).
:::

@@

By summing over $\sigma$ (resp. $v$), these give [(M'5)](mp5).

@@

We leave to the reader the task of translating axioms [(AM1)](am1), [(AM3)](am3), and [(AM4)](am4), which remain to be imposed, into this language (cf. [(1.14)](#chunk-1.14)).
::::

@@

- In [(AM3)](am3), we ask for each $\QQ_\ell$-sheaf $M_\ell$ induced by $M_\AAf$ to come from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$.
- In [(M'4)E](#Mp4E), for each residue field $k(v)$ of $S$, we ask for $M_\crisv$ over the field of fractions $K_v$ of Witt vectors over $k(v)$, endowed with a semi-linear $\phi_v$.
  The crystalline comparison isomorphism of [(M'5)E](#Mp5E) then becomes
  $$
    M_\DR\otimes E_v \simto M_\crisv\otimes_{K_v}E_v.
  $$
:::

@@

The axioms are modified as follows.
In [(AM1)](am1), we want a variation of mixed Hodge structures;
[(AM3)](am3) becomes: $M_\ell$ comes from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$;
for [(AM4)](am4), a condition is imposed for every closed point of $S$.

@@

- [(M1)](m1) and [(M7)](m7): $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $F_\infty=-1$.
- [(M'1)](mp1): $\ZZ(1)_\BB(C)=2\pi i\ZZ\subset C$.
- [(M2)](m2) and [(M'2)](mp2): $\ZZ(1)_\DR=\QQ(1)_\DR=\QQ$.
- [(M3)](m3) and [(M9)](m9): $\ZZ(1)_{\ZZ/n}$ is the group $\mu_n\subset\CC^\times$ of $n$-th roots of unity.
  The transition morphisms are the $\mu_n\to\mu_m\colon x\mapsto x^{n/m}$ for $m\mid n$.
  The action of $\Gal(\bar{\QQ}/\QQ)$ is its action on the $\mu_n$.
  In the variant [(M'3)](mp3), we have $\ZZ(1)_{\ZZ/n}(\bar{k})=\mu_n(\bar{k})$.

  We thus have that $\ZZ(1)_\ell=\limproj\mu_{\ell^n}(\CC)$.
- [(M4)](m4), [(M10)](m10), and [(M'4)](mp4): $\ZZ(1)_\crisp=\QQ_p$ and $\phi_p=1/p$.
- [(M5)](m5): The canonical comparison isomorphism $\comp_{\DR,\BB}$ is induced by the inclusions $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $\ZZ(1)_\DR=\QQ\subset\CC$, i.e.
  $$
    \ZZ(1)_\BB\otimes\CC \simto \CC \simfrom \ZZ(1)_\DR\otimes\CC.
  $$
  The isomorphism $\comp_{\ZZ/n,\BB}$ [(1.25)](#chunk-1.25) is induced by
  $$
    \exp(x/n)\colon \ZZ(1)_\BB \to \ZZ/n(1)\subset\CC^\times.
  $$
  The isomorphism $\comp_{\crisp,\DR}$ is induced by the inclusion $\QQ\subset\QQ_p$.
- [(M'5)](mp5): Replace $\CC$ by $C$ in [(M5)](m5).
- [(M6)](m6) and [(M8)](m8): $\ZZ(1)$ is of pure weight $-2$, and $\ZZ(1)_\DR$ is of pure Hodge filtration $-1$: the Hodge type is $(-1,-1)$.
:::

@@

::: {.trmenv #M7P title="(M7)P" latex="{(M7)P}"}
\oldpage{108}The data of $F_\infty$ for $E$ is equivalent to that of an involution of $P_\BB$ that is compatible with that of $M_\BB$.
Axiom [(AM2)](am2) for $E$ is that this involution is induced by the $\CC$-antilinear involution of $P_{\BB,\CC}=\PP_{\DR,\CC}$ with fixed points $P_{\DR,\RR}$.
:::

::: {.trmenv #M8P title="(M8)P" latex="{(M8)P}"}
The data of the Hodge filtration of $E_\DR$ is equivalent to that of the intersection $F^0P_\DR$ of $F^0E_\DR$ over $P_\DR$, which is a torsor for $F^0M_\DR$: for $p>0$, $F^pE_\DR=F^pM_\DR$; for $p=0$, $F^0E_\DR$ is generated by $F^0P_\DR$; and for $p<0$, $F^pE_\DR$ is the sum of $F^0E_\DR$ and $F^pM_\DR$.
In other words, the data [(M8)](m8) for $E_\DR$ is that of a reduction of the structure group of $P_\DR$, from $M_\DR$ to $F^0M_\DR$.

The data of [(M1)](m1), [(M2)](m2), $\comp_{\DR/\BB}$, and [(M8)](m8) are equivalent to the data of a torsor $F^0P_\DR$ for $F^0M_\DR$ and, in the $M_\DR\otimes\CC=M_\BB\otimes\CC$ torsor $P_{\DR,\CC}$ which is induced by $F^0M_\DR\to M_\DR\otimes\CC$, a $\QQ$-structure $P_\BB\subset P_{\DR,\CC}$ that is an $M_\BB$ torsor.
:::

@@

We leave to the reader the task of translating axioms [(AM1)](am1) to [(AM5)](am5) for $P$.
We note only that [(AM1)](am1) and [(AM4)](am4) follow from the same axioms for $M$, and that [(AM2)](am2) and [(AM5)](am5) can be expressed at best as functoriality in $\CC$, cf. [(1.14)](#chunk-1.14).
::::
